---
title: "Four Statistical Tests"
author: John Ashley Burgoyne
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(effsize)
knitr::opts_chunk$set(echo = FALSE)
billboard <- aenoa::billboard
billboard_plus <- 
    billboard %>% mutate(tempo_beats = tempo * duration_seconds / 60)
```

## Packages and Data

### Packages

This tutorial uses the following packages:

* **effsize** for computing Cohen's *d*
* **tidyverse** for general data processing

### Data

This tutorial has a data files pre-loaded as the variable `billboard`. The data
are a subset of the 
[McGill Billboard data set](http://ddmal.music.mcgill.ca/research/billboard).
This data set contains a random sample of popular songs from the *Billboard* Hot
100 chart for popular music in the U.S. For each song, this tutorial subset
contains eight columns.

| Column Name | Description  |
|---|------------------------------------------|
| **year** | Year the song was released |
| **tonic** | Letter name of the tonic key of the song |
| **mode** | Mode of the tonic key of the song (major or minor) |
| **tempo** | Tempo of the song in beats per minute |
| **loudness** | Loudness of the song (in decibels relative to full scale [dBFS]; –96 dbFS is the softest possible song and 0 dbFS is the loudest possible) |
| **danceability** | Spotify’s prediction of how ‘danceable’ the song is (0 is the least danceable and 1 is the most danceable) |
| **duration_seconds** | Length of the song in seconds |
| **duration_beats** | Length of the song in musical beats |

Use the box below to view the raw data.

```{r bboard, exercise = TRUE}
# Enter your code here.
```

```{r bboard-solution}
billboard
```

## Descriptive Statistics

Before beginning with statistical tests, it is important to get a sense of the 
data through plots and basic descriptive statistics.

### Categorical (Nominal) Variables

We use bar charts to visualise nominal variables. The following code chunk 
displays a bar chart for `mode`. Revise it to make one for `tonic`. What is the
mode of the distribution of tonics?

```{r geom-bar, exercise = TRUE}
ggplot(billboard, aes(x = mode)) + 
    geom_bar() +
    labs(x = "Mode", y = "Count")
```

### Continuous Variables

We use histograms to visualise continuous variables. The following code chunk
displays a bar chart for `duration_seconds`.

Revise it to make one for `danceability`.

```{r geom-histogram, exercise = TRUE}
ggplot(billboard, aes(x = duration_seconds)) +
    geom_histogram(binwidth = 30) +
    labs(x = "Duration (s)", y = "Count")
```

```{r skew}
quiz(
    question(
        "Is danceability positively or negatively skewed (the third moment)?",
        answer("Positively"),
        answer("Negatively", correct = TRUE),
        answer("Neither"),
        allow_retry = TRUE             
    )
)
```

You may also wish to explore some of the other continuous variables in the code
box above.

The following code box computes the means of the continuous variables, one of
the measures of central tendency (i.e., the first moment). Modify it to compute
a measure of dispersion (i.e., the second moment).

```{r moments, exercise = TRUE}
billboard %>% 
    summarize(
        mean_tempo = mean(tempo),
        mean_loudness = mean(loudness),
        mean_danceability = mean(danceability),
        mean_duration_s = mean(duration_seconds),
        mean_duration_b = mean(duration_beats)
    )
```

## Categorical-Variable Tests

There are two important categorical variables for these data: `tonic` and 
`mode`.

### Chi-Square Test

To see whether there is a relationship between them, we need to use a
$\chi$~2~ test:

```{r chisq, exercise = TRUE}
chisq.test(billboard$tonic, billboard$mode)
```

We would report the results as follows:

> There was a significant relationship between tonic and mode, $\chi$~2~(11) = 106, *p* < .001.

What is the null hypothesis for this test?

### Effect Size: Odds Ratio

Every statistical test is better when we report a standard effect size. Andy
Field doesn't provide good code for the effect size for $\chi$~2~ tests, but
here is a template that you can use in your own work.

```{r contingency, exercise = TRUE}
billboard %>% 
    group_by(tonic) %>% 
    count(mode) %>% 
    spread(mode, n) %>% 
    ungroup %>% 
    mutate(
        major_odds = major / (sum(major) - major),
        minor_odds = minor / (sum(minor) - minor),
        odds_ratio = minor_odds / major_odds)
```

## Continuous-Variable Tests

This data set contains many possible pairs of continuous variables.

### Scatter plot

The following code produces a scatter plot of tempo vs. duration in seconds.
Modify it to show a scatter plot of tempo vs. danceability.

```{r tempo-duration-plot, exercise = TRUE}
ggplot(billboard, aes(x = tempo, y = duration_seconds)) +
    geom_point() +
    geom_smooth() +
    labs(x = "Tempo (BPM)", y = "Duration (s)")
```

```{r tempo-duration-quiz}
quiz(
    question(
        "Which of the following tempi does Spotify rate the most danceable?",
        answer("80"),
        answer("120", correct = TRUE),
        answer("180"),
        allow_retry = TRUE
    )
)
```

### Correlation Test

For these pairs, we want to test whether their correlations are significant.
For example, is there a significant correlation between tempo and the length of
a song in seconds?

```{r tempo-duration, exercise = TRUE}
cor.test(billboard$tempo, billboard$duration_seconds)
```

We would report the results of this test as follows.

> There was no significant relationship between tempo and duration in seconds, *t*(886) = --1.2, *p* = .25, *r* = --.04.

Use the box above to check for significant correlations between tempo and danceability.

```{r tempo-danceability}
quiz(
    question(
        "Are tempo and danceability significantly correlated (alpha = .05)?",
        answer("Yes", correct = TRUE),
        answer("No")
    )
)
```

Do these results correspond to what you expected from the scatter plot? Why or 
why not?

### Effect Size: Correlation Coefficient

Correlation coefficients are already standardised effect sizes -- do you 
remember the cut-offs for small, medium, and large effects? -- and so we do not
need to compute anything extra for a correlation test.

### Dependent (Paired) *t* Tests

If there are two continuous variables on the same scale, we can supplement a
correlation test with a paired (dependent) *t* test. For example, Spotify 
estimates the duration of a song in beats separately from its estimate of the
tempo of a song (in beats per minute). Because we have the duration of every
song in seconds, we can compute how many beats long a
song 'should' be, based on Spotify's tempo estimate. Let's call that estimate
`tempo_beats`.

```{r tempo-beats, echo = TRUE}
billboard_plus <-
    billboard %>% 
    mutate(tempo_beats = tempo * duration_seconds / 60)
billboard_plus
```

Now, `tempo_beats` and `duration_beats` are on the same scale (i.e., they use 
the same unit of measurement, namely, musical beats). 

A scatter plot shows an interesting relationship between these two variables.
(Bonus question: Can you think of a reason why we might see this pattern?)

```{r tempo-plot, echo = TRUE}
ggplot(billboard_plus, aes(x = duration_beats, y = tempo_beats)) +
    geom_point() +
    labs(x = "Actual Duration (beats)", y = "Theoretical Duration (beats)")
```

As we would expect, the two variables are highly correlated.

```{r dep-cor, echo = TRUE}
cor.test(billboard_plus$duration_beats, billboard_plus$tempo_beats)
```

But in addition to the correlation, we might also be interested in whether the
the theoretical duration in beats (`tempo_beats`) is systematically higher or
lower than the actual duration in beats. 

```{r dep-means, echo = TRUE}
billboard_plus %>% 
    summarise(
        mean_duration_beats = mean(duration_beats),
        mean_tempo_beats = mean(tempo_beats))
```

This comparision is a difference in
two means, and so we need a *t* test. Because our variables are paired, however,
we need to use the paired (dependent) *t* test.

```{r dep-t, echo = TRUE}
t.test(billboard_plus$duration_beats, billboard_plus$tempo_beats, paired = TRUE)
```

We also need to compute Cohen's *d* (see next section).

```{r dep-d, echo = TRUE}
cohen.d(billboard_plus$duration_beats, billboard_plus$tempo_beats)
```

We would report all of these tests together as follows:

> The theoretical duration of songs in beats, based on Spotify's tempo estimate, and the actual duration in beats were highly correlated, *t*(886) = 25, *p* < .001, *r* = .64, but the tempo-based estimate of duration in beats was systematically higher than the actual duration in beats, *M* = 36 beats, *t*(887) = 7.7, *p* < .001, *d* = .22.

## Mixed-Variable Tests

Of the two categorical variables in these data, one has two categories (i.e., 
is binary) and one has multiple categories.

### Box plots

The following code generate a box plot illustrating the distribution of song
durations (in seconds) versus mode. Modify it to make a box plot showing the
distributions of danceability for each mode. Then modify it again to show the
distributions of danceability for each tonic.

```{r boxes, exercise = TRUE}
ggplot(billboard, aes(x = tonic, y = duration_seconds)) +
    geom_boxplot() +
    labs(x = "Mode", y = "Duration (s)")
```

### Independent *t* Tests

The following code conducts a *t* test for the difference in duration (in
seconds) between modes. 

```{r t-ind, exercise = TRUE}
t.test(billboard$duration_seconds ~ billboard$mode)
```

We also need to compute Cohen's *d* as a standardised effect size for a *t*
test.

```{r d-ind, exercise = TRUE}
cohen.d(billboard$duration_seconds ~ billboard$mode)
```

We would report the results of the test as follows:

> There was a small but significant effect of mode (major vs. minor) on song duration in seconds, *t*(323) = 3.6, *p* < .001, *d* = .32.

Modify the code blocks above to conduct a *t* test for the difference in 
danceability between modes.

```{r d-size-quiz}
quiz(
    question(
        "Although the point estimate of Cohen's *d* represents a small effect of mode on danceablity, based on the 95-percent confidence interval, what is the largest possible effect size we could consider?",
        answer("Negligible"),
        answer("Small"),
        answer("Medium", correct = TRUE),
        answer("Large"),
        incorrect = "Incorrect. Remember, for Cohen's *d*, a small effect is beyond 0.2, a medium-sized effect is beyond 0.5, and a large effect is beyond 0.8.",
        allow_retry = TRUE
    )
)
```

### ANOVA

The following tests conducts an ANOVA on differences in average duration (in
seconds) for different tonics.

```{r anova, exercise = TRUE}
oneway.test(billboard$duration_seconds ~ billboard$tonic)
```

We would report the results of the test like this:

> ANOVA showed no significant difference in average song duration across
tonics, *F*(11, 259) = 1.3, *p* = .20.

Because the result is not statistically significant, usually we would stop here,
but if the result had been significant, the next step would to look at *t* tests
for every possible pair of tonics. When we conduct multiple *t* tests at the
same time, it is important to correct for making multiple comparisions:
otherwise our changes of making a Type I error are much larger than we 
planned! Let's check the tests for this ANOVA.

```{r post-hoc, exercise = TRUE}
pairwise.t.test(billboard$duration_seconds, billboard$tonic)
```

None of the pairs shows a significant result, which is not surprising: that
is why our original `oneway.test` was not significant.

Modify the above code chunks to test differences in loudness across modes.

```{r anova-quiz}
quiz(
    question(
        "Which of the following pairs of tonics shows a significant 
        difference in loudness?",
        answer("E and F-sharp"),
        answer("B and E", correct = TRUE),
        answer("B and F"),
        answer("F and F-sharp"),
        random_answer_order = TRUE,
        allow_retry = TRUE
    )
)
```

